---
title: "Week 02"
author: "Matias Andina"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
    keep_md: true
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
theme_set(theme_bw())
```

Let's load the data for this week:

```{r load-data , message=FALSE, warning=FALSE}
tuesdata <- tidytuesdayR::tt_load(2021, week = 2)
transit_cost <- tuesdata$transit_cost
```

Let's take a look at the data:

```{r mini-inspect}
head(transit_cost)
```

Let's perform some inspection of the values

```{r inspection, message=FALSE, warning=FALSE}
psych::describe(transit_cost)
```

Check if we have `NA`s and where.

```{r mo-inspection}
transit_cost %>% is.na() %>% colSums() %>% enframe(value = "NA_count")
```

### Outlier inspection

Quick check for extreme values

```{r length, message=FALSE, warning=FALSE}
cowplot::plot_grid(
ggplot(transit_cost, aes(length)) +
  geom_histogram(),
ggplot(filter(transit_cost, length<5000),
       aes(length)) +
  geom_histogram(color="black")
)
```

Remove outliers in `length`

```{r remove-length}
transit_cost <- filter(transit_cost, length < 5000)

```

### Cleaning year variables

The variables `start_year` and `end year` are not a `double` and are contaminated. We can also make a `duration` variable, so that we account for the duration of the project. The ones with `4 years` will be counted as `duration`.

```{r inspect-years}
unique(transit_cost$start_year)
unique(transit_cost$end_year)
unique(transit_cost$year)
```

```{r clean-years, message=FALSE, warning=FALSE}
transit_cost <- transit_cost %>% 
  # duration and cleaning of years
  # NAs by coercion
  mutate(duration = ifelse(str_detect(start_year, "years"),
                           parse_number(start_year),
                           as.numeric(end_year) - as.numeric(start_year)),
         start_year = as.numeric(start_year),
         end_year = as.numeric(end_year))
```

Let's inspect the new `duration` variable:

```{r duration-hist, message=FALSE, warning=FALSE}
transit_cost %>% 
  ggplot(aes(duration))+
  geom_bar(color="black")
```

It looks like it takes 5 years to finish a urban rail project. It's also independent of the `length`, which is super interesting!

```{r duration-hist-fill, message=FALSE, warning=FALSE}
transit_cost %>% 
  mutate(cutt = cut(length, seq(0, 200, 40))) %>% 
  #filter(!is.na(duration)) %>%
  group_by(duration, cutt) %>% 
  count() %>% 
  ggplot(aes(duration, y=n, fill=cutt)) +
  geom_col() +
  labs(title="It takes roughly 5 years to complete a urban rail",
       fill="Project length (Km)",
       x="Project Duration (years)", y="Count",
       caption="Source: Global Transit Costs | Viz: @NeuroMLA") +
  ggthemes::theme_clean()+ 
  scale_fill_brewer(palette = "Set1")+
  theme(legend.position="bottom")
```

```{r, warning=FALSE}
ggplot(transit_cost,
       aes(length, cost_km_millions)) +
  geom_point() +
  ggrepel::geom_text_repel(data = filter(transit_cost, cost_km_millions > 1000), 
            aes(length, cost_km_millions, label=city),
            direction="x", 
            segment.color = "black") +
  ggrepel::geom_text_repel(data = filter(transit_cost, length > 100), 
                           aes(length, cost_km_millions, label=city),
                           direction="y", 
                           segment.color = "black")+
  labs(title="Urban rails of the World: Paris beats NYC",
       x="Length (Km)",
       y="Cost per Km in millions",
       caption="Source: Global Transit Costs | Viz: @NeuroMLA")

```

### Stations

Inspection the number of stations

```{r station-hist, warning=FALSE, message=FALSE}
ggplot(transit_cost, aes(stations)) +
  geom_histogram(color="black")
```
Something interesting to check is whether the stations are somewhat evenly distributed across the length of the project.

```{r station-length-hist, warning=FALSE, message=FALSE}
ggplot(transit_cost, aes(stations/length)) +
  geom_histogram(color="black", binwidth = 0.1) +
  geom_density(aes(y = ..density..*(nrow(transit_cost) * 0.1)))+
  labs(
    title = "Station distribution across urban rails in the world",
    subtitle=glue::glue("Average of {round(mean(transit_cost$stations/transit_cost$length, na.rm=T), 2)} stations per km"))
```
Interestingly, urban rails that are very long, also keep the `stations/length` ratio around the mean.

```{r stations-length-long, warning=FALSE}
ggplot(transit_cost, aes(stations/length, length)) +
  geom_point()
```


Since I saw this dataset contained `start_year` and `end_year` I was dying to make this plot. Wish fulfilled!

```{r}
transit_cost %>% 
  select(city, start_year, end_year, line) %>% 
  mutate(line = factor(paste(city, line)),
         line = fct_reorder(line, desc(end_year))) %>% 
  filter(complete.cases(.)) %>% 
  ggplot(aes(y=line)) + 
  geom_segment(aes(x=start_year, xend=end_year, yend=line, color = line))+
  theme(legend.position="none",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())+
  labs(title = "Project urban rail: Duration around the World",
       subtitle = "Each segment represents one urban rail line construction period (start to end years)",
       x = "Year",
       y = "urban rail line",
       caption = "Source: Global Transit Costs | Viz: @NeuroMLA")+
  theme(panel.background = element_rect(fill="gray90"),
        panel.grid = element_blank(), panel.grid.major.x = element_line(color="gray50",linetype = 2))+
  #scale_color_viridis_d(option = "D")+
  NULL
```

Seeing how a chunk of lines get completed at the same time, I got superstitious and decided to check whether projects start or end on even/odd years. It looks like if you are going to dig, you are more likely to choose an even year, and you will finish an an odd year 🤔

```{r}
transit_cost %>% 
  count(start = start_year %% 2,
        end = end_year %% 2) %>%
  mutate_at(vars(c("start", "end")),
            ~ifelse(. == 0, "even", "odd")) %>% 
  filter(complete.cases(.)) %>% 
  ggplot()+
  geom_point(aes(start, end, size =n ,color=n))+
  labs(title = "There's something odd going on",
       subtitle = "Most urban rails start on odd years and finish on even years",
       x = "Project start year",
       y = "Project end year",
       caption = "Source: Global Transit Costs | Viz: @NeuroMLA")+
  theme(legend.position = "bottom")

```
